Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-09-19 manuscript/paper.tex
intro ... many references...

We consider
rigid rotation and tilting of the 
%BO$_{6/2}$ 
octahedra
 in ABO$_3$ perovskites
and study the
effect of the crystal field of the nearest A and B ion cages 
%that immediately surround the octahedra
on the single electron spectrum of the 
central 
transition metal ions.
While the crystal field of the octahedra 
that 
% quenches the orbital angular momentum of the central B ion and 
creates $e_g$ and $t_{2g}$ manifolds
do not change with such rigid rotations, 
the field of 
%A and B ionic cages %that immediately surround the octahedra
A and B cages
deforms significantly %with the octahedral rotations 
from the ideal octahedral field
in a way that
leads to sizeable splittings in the $e_g$ and $t_{2g}$ manifolds of the central B ion.
%Similar to the Jahn-Teller problem,
The lowering of the ground state energy
of a given many-electron configuration
due to these splittings
can thus be an important driving force behind the octahedral rotations 
and associated structural transformations.
We find that the size of the splitting and the local orbital ordering are
determined by a competition between the A and B cages, 
as well as a competition between the tilt and rotation angles.

@@

%Their crystal structures 
The perovskite structures
have a common formula unit ABX$_3$,
where 
%positive ions A and negative ions X form a cube and an octahedron around positive ion B.
A and B are positive ions while X are negative ions 
%whose six-fold coordination forms octahedra around B.
that form octahedra around  B.
In a cubic perovskite,
the octahedra are symmetric, 
but they can distort or simply rotate to lower the symmetry
and produce a tetragonal or orthorhombic structure.
While the tetragonal distortion of the octahedra is associated with
a (cooperative) Jahn-Teller effect
(i.e., the orbital degeneracy of the metal ion coordinated by the anionic octahedron
is lifted, which stabilises the distortion),
%[as it distorts the octahedral field at the central site that split its electronic states],
the rotation and tilting of the octahedra are almost always 
explained empirically in terms of 
the relative sizes of the three ions
with the
Goldschmidt tolerance factor. 
%[probably because rigid rotation of the octahedra do not change its potential at the central site]
%We only find two exceptions.

%\az{terminology: tilting, rotation.... vague is ok here...}

Based on density functional theory (DFT) calculations,
a few recent works
~\cite{garcia-fernandezJPCL10,CammarataJCP14,LeeCM16,YoshidaPRL21}, 
however, 
suggest that the octahedral rotation and tilting is 
a second order Jahn-Teller effect
%\az{describe SOJT in a sentence: coupling to the low lying unoccupied states on the JT ion, or orbital hybridisation with other ions, coupling between bands...}
(i.e., the distortion is stabilised by a coupling between the ground and excited adiabatic potential energy surfaces).
% garcia-fernandezJPCL2010
Garcia-Fernandez~et~al.
study
potassium fluoride perovskite family KMF$_3$ (M$=$Ca$^{2+}$ or a $3$d transition metal)
and found that the tilting angle
depends on the covalent bonding through the mixing of occupied and 
low-lying unoccupied orbitals of the transition metal~\cite{garcia-fernandezJPCL10}.
% CammarataJCP2014
Cammarata and Rondinelli
find a negative correlation between metal-oxygen bond covalency 
and octahedral rotation in orthorhombic perovskites oxides~\cite{CammarataJCP14}.
% Lee, nick2016
Lee~et~al. find 
similar results for metal-iodine bonding in
lead iodide perovskites,
where the role of 
the hydrogen bonding 
in a hybrid perovskite with
organic ammonium ions at A sites
is also found to be crucial~\cite{LeeCM16}.
Yoshida~et~al.
use DFT along with 
representation theory analysis 
and 
indicate that 
the rotations are induced by bonding interaction 
between the valence and conduction band Bloch states at different wavevectors
that are mapped onto each other on Brillouin zone folding
 for the associated lattice distortion~\cite{YoshidaPRL21}.

\begin{figure}[htbp]
\begin{center}
%\includegraphics[width=1\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/struc/structures-combined-labelled-1}
\includegraphics[width=1\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/struc/structures-combined-labelled-2}
%\includegraphics[height=5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cell-cubic}
%\includegraphics[height=5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cell-pbnm-15-15}\\
%\includegraphics[height=4.5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cages-cubic-z}
%\includegraphics[height=4.5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cages-rotated-15-15-z}
%\includegraphics[height=6.cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/AB-cages-deformed}
\caption{
Crystal structures and coordination cages.
%Crystal structure of cubic and orthorhombic perovskites (a-b) and ion coordination cages around B ions (c-d).
Crystal structures of cubic (a) and orthorhombic (b) perovskites.
The anions (red) form the shown octahedra around 
the B cations (gold), while the A cations (green) sit in the voids between the octahedra.
In the orthorhombic structure shown here (b),
 the octahedra are rotated and tilted by $15^\circ$.
(c-d)
The coordination cages of A %(green) 
and B 
%(golden) 
cations just next to the anion 
%(red) 
octahedron
around a B ion in cubic %(c)
 and orthorhombic %(d)
  perovskites.
In the anion octahedron's frame,
the cages are symmetric and aligned to the octahedron in the cubic perovskites (c),
while they are rotated and slightly distorted in the orthorhombic perovskites (d).
}
\label{fig:struc}
\end{center}
\end{figure}

We take a completely different approach to this problem.
We go back to the basics
and see 
what is it that lifts 
 the orbital degeneracy in the classical JT effect?
It is a specific 
deformation of the crystal field 
%(${\sim z^2}$)
at the transition metal (JT ion) site.
Does the same happen in the case of octahedral tilting and rotations?
While the potential of an octahedral anion cage 
in its own frame
does not change 
%in case of 
with
rigid rotations for obvious reason,
the total crystal field 
does deform indeed
because
 the rest of the lattice is not only rotated
  in a given octahedron's frame (that turns its non-spherical potential)
 but also distorted owing to the accompanying structural transformation.
%So, in principle, there should be a (first order) JT effect at play.
Using this insight
and elements of crystal field theory,
here in this manuscript,
we 
consider next two coordination cages around an octahedron
% explore this problem in detail
and find that,
just like a first order JT effect,
 the deformed crystal field 
in this case also lifts the orbital degeneracy,
by a significant amount for typical system parameters,
signalling the instability against the rotations.

@@

Since the octahedra are connected to each other 
(making a 3-dimensional network in the perovskite structure),
rotating one octahedron rotates all other in the crystal in a consistent fashion;
 see Figs.~\ref{fig:struc}(a,b).
We consider an octahedral \emph{rotation} $\beta$ around $001$ direction
(that transforms the structure from cubic Pm$\bar 3$m to tetragonal P4/mbm)
%({No.~$127$})
followed by a \emph{tilting} $\alpha$ around the new $110$ direction
(transforming to orthorhombic Pbnm, 
%({No.~$62$})
or Imma 
%({No.~$74$})
 at ${\beta=0}$).
We use the rotation and tilt terminology for these 
two angles throughout the rest of the manuscript.

@@

\subsection{Atom/ion in an octahedral field: $t_{2g}$  and $e_g$ manifolds}
\label{sec:egt2g}
Consider six O ions each of charge $-q_o$ making an octahedron of size $a$ (placed at the face centres of a cube of edge length $a$).
The non-spherical part of the potential they
 produce at the centre of the octahedron
 %that is 
 relevant for the d-orbitals contains 
 only two $l=4$ components
  in a specific ratio (${V_{4,0}:V_{4,4}=1:\sqrt{5/7}}$)
 and is called octahedral field. 
It is given by~\cite{pavariniChap},
\begin{align}
\label{eq:voct}
V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right].
\end{align}
%where the monopole and $l>4$ components have been dropped
The crystal field Hamiltonian of a transition metal ion B 
%at the centre 
in the field $V_{oct}(\vec r)$
turns out to be diagonal,
with its states split into two manifolds $\Delta$ apart:
$H_{m,m} = -\Delta/3$ for $m\in\{-2,-1,1\}$ ($xy,yz,zx$ orbitals)
called $t_{2g}$,
and 
$H_{m,m} = +2\Delta/3$ for $m\in\{2,0\}$ ($x^2-y^2,z^2$ orbitals)
called $e_g$.
The crystal field splitting between the two manifolds is
${\Delta=\frac{160}{3}\frac{q_o}{a} D_{4}}$.

@@

The Jahn--Teller effect~\cite{Jahn-Teller} describes the 
geometric and electronic instability of a %non-linear 
molecule, cluster or crystal with
a symmetric structure and orbitally degenerate ground state
%(i.e., except Kramers degeneracy)
against structural distortions.
The distortions  
reduce the symmetry of the structure
and lift the electronic degeneracy.
This lowers the energy of the system proportional to
the distortion.
The elastic energy of the structure that increases as the square of the distortion
competes with this linear term and the optimum structure 
is obtained at a finite distortion.

@@

The dependence of the elastic energy on the distortion can be roughly explained
as follows.
Consider an 
elongation or a compression of an octahedron along one of its axes by $a\delta$,
where $a$ is the lattice constant of the cubic structure.
The corresponding change in the unit cell volume would be linear in $\delta$
as $\sim a^3 \delta$
so the elastic energy
would change as $\sim\delta^2$.
Assuming a linear change 
 in the electronic spectrum,
  the total energy has a form
$E=-\rho\delta + K \delta^2$, where $\rho$ and $K$ are constants 
describing the changes in the electronic and elastic energies,
and would be optimum at $\delta=\rho/2K$.
In general, as long as the two competing terms 
are not the same degree monomial in the distortion,
the ground state can have a finite distortion.
We will use this argument and focus only on the changes
 in the electronic energy throughout this paper.

@@

\com{
At $\delta=0$, keeping only the non-spherical components, 
we obtain the
 octahedral potential for a symmetric octahedron~\cite{pavariniChap},
\begin{align}
\label{eq:voct}
\lim_{\delta\to0}V^{JT}(\vec r) \to V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right],
\end{align}
}

@@

In a crystal, obviously, not only the nearest coordination cage of A,B ions but also
 the other ions (including O) at longer distances
  participate in the effect considered in this paper.
To see how much our results change when 
%the potential of 
 the full lattice (FL) is considered instead of just the nearest AB cages,
we numerically calculate the multipole components of the potential
 of the full lattice by following the recipe in Ref.~\cite{FinnisPRL98,PaxtonPRB08,PaxtonNotes}
and Ewald summation method~\cite{Ewald}.
To do this, 
we borrow relevant routines from the tight binding package in Questaal suite~\cite{Questaal} that implements these methods.

@@

We find that the
potential of the FL does not change the qualitative picture we presented
but merely shifts the balance between A and B cages towards smaller $q_A$.
This can be seen in Fig.~\ref{fig:t2g-lattice}
where the energies of the $t_{2g}$ states as a function of $q_A/q_o$
are compared for the two cases
at $\alpha=0$, $\beta=20^\circ$ (and $\Delta=4q_o/15a$).
In case of AB cages, the crossing of the 
($xy$ and $yz/zx$)
energy levels occurs 
at $q_A/q_o\simeq1.1$,
whereas the FL moves it to $q_A/q_o\simeq0.3$.
The slopes of the lines are almost the same in both cases,
meaning that the splitting due to octahedral rotations 
would be the same in both cases if $q_A$ is measured 
from their crossing points.
The shift towards smaller $q_A$ in case of FL
can be explained by noting that
the next coordination cage beyond the first AB cages
has O ions
(at distances $\sqrt{5}a/2$ at $\alpha,\beta=0$) %and $\sqrt{5/2}a$ at $\alpha,\beta=0$),....
%\az{both of which supports the A cage potential.??? plot and see... }
that supports A cage
against the B cage, lowering the value of $q_A$ that balances it
at the crossing in Fig.~\ref{fig:t2g-lattice}.
It increases the agreement between the two cases presented significantly,
by moving the crossing point 
to $q_A/q_o\simeq0.8$ (not shown in Fig.~\ref{fig:t2g-lattice}).

@@

\label{Bibliography}
\bibliographystyle{apsrev4-1}
\bibliography{perovskites}

>>>>

We consider
rigid rotation and tilting of the 
%BO$_{6/2}$ 
octahedra
 in ABO$_3$ perovskites
 from the viewpoint of Jahn-Teller problem
 and study the
effect of the crystal field of the nearest A and B ion cages 
%that immediately surround the octahedra
on the single electron spectrum of the 
central 
transition metal ions.
While the crystal field of the octahedra 
that 
% quenches the orbital angular momentum of the central B ion and 
creates $e_g$ and $t_{2g}$ manifolds
do not change with such rigid rotations, 
the field of 
%A and B ionic cages %that immediately surround the octahedra
A and B cages
deforms significantly %with the octahedral rotations 
from the ideal octahedral field
in a way that
leads to sizeable splittings in the $e_g$ and $t_{2g}$ manifolds of the central B ion.
%Similar to the Jahn-Teller problem,
The lowering of the ground state energy
of a given many-electron configuration
due to these splittings
can thus be an important driving force behind the octahedral rotations 
and associated structural transformations.
We find that the size of the splitting and the local orbital ordering are
determined by a competition between the A and B cages, 
as well as a competition between the tilt and rotation angles.

@@

%Their crystal structures 
The perovskite structures
have a common formula unit ABX$_3$,
where 
%positive ions A and negative ions X form a cube and an octahedron around positive ion B.
A and B are positive ions while X are negative ions 
%whose six-fold coordination forms octahedra around B.
that form octahedra around  B.
In a cubic perovskite,
the octahedra are symmetric, 
but they can 
distort~\cite{
BersukerCCR75,
KugelSPU82,
MillisN98,
TerakuraPMS07,
HalcrowRSC13,
LufasoAC04}
or simply 
rotate~\cite{
GlazerAC72, ThomasAC96,
WoodwardAC97a,WoodwardAC97b,
HowardAC98,StokesAC02,AngelPRL05}
 to lower the symmetry
and produce, e.g., a rhombohedral, tetragonal or orthorhombic structure.
% cubic perovskite [spacegroup 221] can transform to:
% crystal system: spacegroups
% triclinic: 2
% monoclinic: 11,12,15
% orthorhombic: 59,62,63,71,74
% tetragonal 127,139,140
% rhombohedral/trigonal: 167
% cubic: 204, 221 
Octahedral rotations also occur in other perovskite related structures
and
have been extensively 
studied
due to their crucial role in determining various physical properties,
e.g., in 
ferroelectric~\cite{GhosezNM11,
SchickPRB14,
SimPRB14,
LeePRL17,
WangPRL22,
ZhouPRB22,
GaoPRB23} and 
multiferroic~\cite{BenedekPRL11,
YangPRL14,
YePRB18,
ZhangPRL20,
ZhangYajunPRL20, 
ZhouPRB21,
ZhangPRL22}
systems.
The interplay of octahedral rotations
and 
spin-orbit coupling~\cite{AmatNL14,KrachJPCM23},
breathing distortions~\cite{BalachandranPRB13},
strain~\cite{Aguado-PuentePRL11},
defects~\cite{JiaPRB22}, and
pressure~\cite{XiangPRB17},
and 
their role in
magnetostructural phase transitions~\cite{JohnsonPRL20},
orbital order~\cite{OgawaPRL12,ChenPRL19}, orbital anisotropy~\cite{HuangPRB21}, 
and
%surface electronic structure under in-plane uniaxial compression~\cite{MoralesPRL23}
surface electronic structure~\cite{MoralesPRL23},
has also been investigated.
% JT interacting with octa tilting:
Since, usually, the octahedra are not just %(Jahn-Teller) 
distorted or rotated but a combination of both,
the influence of these two 
structural mutations %instabilities 
on each other has recently been explored by many authors~\cite{LufasoAC04,CarpenterAC09,ChenPRL19}.

\com{---a cooperative Jahn-Teller 
effect~\cite{JahnPRSL37,JahnPRSL38,
OpikRS57,
Sturge68,
GehringRPP75,
BersukerCCR75,
KugelSPU82,
MillisN98,
LufasoAC04,
TerakuraPMS07,
HalcrowRSC13,
PavariniChap}
whereby the orbital degeneracy of the B ions
is lifted that stabilises the distortion---
}

While the tetragonal distortion of the octahedra is associated with
a (cooperative) Jahn-Teller 
effect~\cite{JahnPRSL37,JahnPRSL38,
OpikRS57,
Sturge68,
GehringRPP75,
BersukerCCR75,
KugelSPU82,
MillisN98,
LufasoAC04,
TerakuraPMS07,
HalcrowRSC13,
PavariniChap}
(i.e., the orbital degeneracy of the metal ion coordinated by the anionic octahedron
is lifted, which stabilises the distortion),
%[as it distorts the octahedral field at the central site that split its electronic states],
the rotation and tilting of the octahedra are almost always 
explained empirically in terms of 
the relative sizes of the three ions
with the
Goldschmidt tolerance factor~\cite{Goldschmidt1926,KieslichCS15}. 
%[probably because rigid rotation of the octahedra do not change its potential at the central site]
%We only find two exceptions.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/struc/structures-combined-bonds-labelled-1}
%\includegraphics[width=1\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/struc/structures-combined-labelled-2}
%\includegraphics[height=5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cell-cubic}
%\includegraphics[height=5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cell-pbnm-15-15}\\
%\includegraphics[height=4.5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cages-cubic-z}
%\includegraphics[height=4.5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cages-rotated-15-15-z}
%\includegraphics[height=6.cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/AB-cages-deformed}
\caption{
Crystal structures and coordination cages.
Crystal structures of cubic (a) and orthorhombic (b) perovskites.
The anions (red) form the shown octahedra around 
the B cations (gold), while the A cations (green) sit in the voids between the octahedra.
In the orthorhombic structure shown here (b),
 the octahedra are rotated and tilted by $15^\circ$.
(c-d)
The coordination cages of A %(green) 
and B 
%(golden) 
cations just next to the anion 
%(red) 
octahedron
around a B ion in cubic %(c)
 and orthorhombic %(d)
  perovskites.
In the anion octahedron's frame,
the cages are symmetric and aligned to the octahedron in the cubic perovskites (c),
while they are rotated and slightly distorted in the orthorhombic perovskites (d).
}
\label{fig:struc}
\end{center}
\end{figure}

Based on density functional theory (DFT) calculations,
a few recent works
~\cite{garcia-fernandezJPCL10,CammarataJCP14,LeeCM16,YoshidaPRL21}, 
however, 
suggest that the octahedral rotation and tilting is 
a second order Jahn-Teller effect
~\cite{HerzbergZPC33,BersukerPL66,KristoffelPSSB67,PearsonJMS83,FultonJCP04,BersukerCR13}
%\az{describe SOJT in a sentence: coupling to the low lying unoccupied states on the JT ion, or orbital hybridisation with other ions, coupling between bands...}
(i.e., the distortion is stabilised by a coupling between the ground and excited adiabatic potential energy surfaces).
% garcia-fernandezJPCL2010
Garcia-Fernandez~et~al.
study
potassium fluoride perovskite family KMF$_3$ (M$=$Ca$^{2+}$ or a $3$d transition metal)
and found that the tilting angle
depends on the covalent bonding through the mixing of occupied and 
low-lying unoccupied orbitals of the transition metal~\cite{garcia-fernandezJPCL10}.
% CammarataJCP2014
Cammarata and Rondinelli
find a negative correlation between metal-oxygen bond covalency 
and octahedral rotation in orthorhombic perovskites oxides~\cite{CammarataJCP14}.
% Lee, nick2016
Lee~et~al. find 
similar results for metal-iodine bonding in
lead iodide perovskites,
where the role of 
the hydrogen bonding 
in a hybrid perovskite with
organic ammonium ions at A sites
is also found to be crucial~\cite{LeeCM16}.
Yoshida~et~al.
use DFT along with 
representation theory analysis 
and 
indicate that 
the rotations are induced by bonding interaction 
between the valence and conduction band Bloch states at different wavevectors
that are mapped onto each other on Brillouin zone folding
 for the associated lattice distortion~\cite{YoshidaPRL21}.

\com{
We take a completely different approach to this problem.
We go back to the basics
and explore the octahedral rotations from the perspective
of (first order) JT effect itself
}

We take a completely different approach to this problem.
We go back to the basics
and see 
what is it that lifts 
 the orbital degeneracy in the classical JT effect?
It is a specific 
deformation of the crystal field 
%(${\sim z^2}$)
at the transition metal (JT ion) site.
Does the same happen in the case of octahedral tilting and rotations?
While the potential of an octahedral anion cage 
in its own frame
does not change 
%in case of 
with
rigid rotations for obvious reason,
the total crystal field 
does deform indeed
because
 the rest of the lattice is not only rotated
  in a given octahedron's frame (that turns its non-spherical potential)
 but also distorted owing to the accompanying structural transformation.
%So, in principle, there should be a (first order) JT effect at play.
Using this insight
and elements of crystal field theory~\cite{BetheAP29,VanVleckPR32,PavariniChap},
here in this manuscript,
we 
consider next two coordination cages around an octahedron
% explore this problem in detail
and find that,
just like a first order JT effect,
 the deformed crystal field 
in this case also lifts the orbital degeneracy,
by a significant amount for typical system parameters,
signalling the instability against the rotations.

@@

Since the octahedra are connected to each other 
(making a 3-dimensional network in the perovskite structure),
rotating one octahedron rotates all other in the crystal in a consistent fashion~\cite{GlazerAC72};
 see Figs.~\ref{fig:struc}(a,b).
We consider an octahedral \emph{rotation} $\beta$ around $001$ direction
(that transforms the structure from cubic Pm$\bar 3$m to tetragonal P4/mbm)
%({No.~$127$})
followed by a \emph{tilting} $\alpha$ around the new $110$ direction
(transforming to orthorhombic Pbnm, 
%({No.~$62$})
or Imma 
%({No.~$74$})
 at ${\beta=0}$).
We use the rotation and tilt terminology for these 
two angles throughout the rest of the manuscript.

@@

\subsection{Atom/ion in an octahedral field: $t_{2g}$  and $e_g$ manifolds}
\label{sec:egt2g}
Consider six O ions each of charge $-q_o$ making an octahedron of size $a$ (placed at the face centres of a cube of edge length $a$).
The non-spherical part of the potential they
 produce at the centre of the octahedron
 %that is 
 relevant for the d-orbitals contains 
 only two $l=4$ components
  in a specific ratio (${V_{4,0}:V_{4,4}=1:\sqrt{5/7}}$)
 and is called octahedral field. 
It is given by~\cite{PavariniChap},
\begin{align}
\label{eq:voct}
V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right].
\end{align}
%where the monopole and $l>4$ components have been dropped
The crystal field Hamiltonian of a transition metal ion B 
%at the centre 
in the field $V_{oct}(\vec r)$
turns out to be diagonal,
with its states split into two manifolds $\Delta$ apart:
$H_{m,m} = -\Delta/3$ for $m\in\{-2,-1,1\}$ ($xy,yz,zx$ orbitals)
called $t_{2g}$,
and 
$H_{m,m} = +2\Delta/3$ for $m\in\{2,0\}$ ($x^2-y^2,z^2$ orbitals)
called $e_g$.
The crystal field splitting between the two manifolds is
${\Delta=\frac{160}{3}\frac{q_o}{a} D_{4}}$.

@@

The Jahn--Teller effect~\cite{JahnPRSL37,JahnPRSL38,OpikRS57,Sturge68,GehringRPP75,KugelSPU82,PavariniChap} describes the 
geometric and electronic instability of a %non-linear 
molecule, cluster or crystal with
a symmetric structure and orbitally degenerate ground state
%(i.e., except Kramers degeneracy)
against structural distortions.
The distortions  
reduce the symmetry of the structure
and lift the electronic degeneracy.
This lowers the energy of the system proportional to
the distortion.
The elastic energy of the structure that increases as the square of the distortion
competes with this linear term and the optimum structure 
is obtained at a finite distortion.

@@

The dependence of the elastic energy on the distortion can be roughly explained
as follows.
Consider an 
elongation or a compression of an octahedron along one of its axes by $a\delta$,
where $a$ is the lattice constant of the cubic structure.
The corresponding change in the unit cell volume would be linear in $\delta$
as $\sim a^3 \delta$
so the elastic energy
would change as $\sim\delta^2$.
Assuming a linear change 
 in the electronic spectrum,
  the total energy has a form
$E=-\rho\delta + K \delta^2$, where $\rho$ and $K$ are positive constants 
describing the changes in the electronic and elastic energies,
and would be optimum at $\delta=\rho/2K$.
In general, as long as the two competing terms 
are not the same degree monomial in the distortion,
the ground state can have a finite distortion.
We will use this argument and focus only on the changes
 in the electronic energy throughout this paper.

@@

\com{
At $\delta=0$, keeping only the non-spherical components, 
we obtain the
 octahedral potential for a symmetric octahedron~\cite{PavariniChap},
\begin{align}
\label{eq:voct}
\lim_{\delta\to0}V^{JT}(\vec r) \to V_{oct}(\vec r) = -\frac{224\sqrt{\pi}}{3}\frac{q_o}{a} \left(\frac{r}{a}\right)^4\left[\ylm{4,0}+\sqrt{\frac{5}{7}}\ylm{4,4} \right],
\end{align}
}

@@

In a crystal, obviously, not only the nearest coordination cage of A,B ions but also
 the other ions (including O) at longer distances
  participate in the effect considered in this paper.
To see how much our results change when 
%the potential of 
 the full lattice (FL) is considered instead of just the nearest AB cages,
we numerically calculate the multipole components of the potential
 of the full lattice by following the recipe in Ref.~\cite{FinnisPRL98,PaxtonPRB08,PaxtonNotes}
and Ewald summation method~\cite{Ewald1921}.
To do this, 
we borrow relevant routines from the tight binding package in Questaal suite~\cite{Questaal} that implements these methods.

@@

We find that the
potential of the FL does not change the qualitative picture we presented
but merely shifts the balance between A and B cages towards smaller $q_A$.
This can be seen in Fig.~\ref{fig:t2g-lattice}
where the energies of the $t_{2g}$ states as a function of $q_A/q_o$
are compared for the two cases
at $\alpha=0$, $\beta=20^\circ$ (and $\Delta=4q_o/15a$).
In case of AB cages, the crossing of the 
($xy$ and $yz/zx$)
energy levels occurs 
at $q_A/q_o\simeq1.1$,
whereas the FL moves it to $q_A/q_o\simeq0.3$.
The slopes of the lines are almost the same in both cases,
meaning that the splitting due to octahedral rotations 
would be the same in both cases if $q_A$ is measured 
from their crossing points.
The shift towards smaller $q_A$ in case of FL
can be explained by noting that
the next coordination cage beyond the first AB cages
has O ions
(at distances $\sqrt{5}a/2$ at $\alpha,\beta=0$) %and $\sqrt{5/2}a$ at $\alpha,\beta=0$),....
%\az{both of which supports the A cage potential.??? plot and see... }
that supports A cage
against the B cage, lowering the value of $q_A$ that balances it
at the crossing in Fig.~\ref{fig:t2g-lattice}.
It significantly increases the agreement between the two cases presented
by moving the crossing point 
to $q_A/q_o\simeq0.8$ (not shown in Fig.~\ref{fig:t2g-lattice}).

@@

\label{Bibliography}
%\bibliographystyle{apsrev4-1}
\bibliography{perovskites}
